Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-05-14 jss/EpiCompare-paper/EpiCompare-paper2.tex
Added slight edits to section 2 (and fixed typos) and added comments for section 3.

{[}EpiCompare emphasizes the value of analyzing epidemics in a
\textit{time-invariant} way - approaches that remove some or all of the
impact of start/end times and recording time scales when preforming the
analysis. In this section we will highlight weaknesses of
time-\textit{de}pendent analysis and define the mathematical
underpinning of the time-invariant approach we take. To accomplish this
goal we will first demonstrate the inability of time-dependent tools to
well quantify a classic epidemic parameter, the reproduction number
\(R_0\). Then we will motivate new time-invariant approaches for more
complex situations where \(R_0\) doesn't capture the complexities of
outbreaks. This will lead us to our final subsection that defines ways
to view epidemics in a time-invariant lense and discusses natural
properties of these representations that allow for clearer ways to
compare models and epidemics.

@@

A lot of epidemiological research is interested in understanding
intrinsic properties of the epidemic, not depending upon exactly when
the epidemic occurred or how frequently data was collected. One of the
most famous numerical summaries of an epidemic is the reproduction
number - \(R_0\), a time-invariant value that is {[}defined as the
expected number of infections caused by a single infect or who is added
to a completely susceptible population.{]} Interestingly,
\citet{Gallagher2020} has shown that the estimation of \(R_0\) {[}can be
sensitive to time-dependent parameters{]} like estimation of {[}the
beginning and end of an epidemic{]}. This time-dependent sensitivity in
estimating \(R_0\) reflects a general problem in epidemiology
surrounding how to transform time-dependent information into insights
into desirable properties of an epidemic.

In many situations, epidemiologists want a much deeper understand of an
epidemic than just the reproducible number. One common time-dependent
tool is a series of time series line plots that track the proportion of
the population in a given state (e.g.~infected) at a given time. Figure
\ref{fig:different-scales-standard} visualizes two different simulated
epidemics with population states (s)usceptible, (i)nfectious and
(r)ecovered. These simulations are actually generated under a discrete
approximation of \citet{Kermack1927}'s SIR model. {[}This model captures
the transitions from one state to the next as a system of ordinary
differential equations, where \(N\) is the total number of individuals,
\(\beta\) is the rate of infection, and \(\gamma\) is the rate of
recovery,{]}

@@

{[}From this model, the reproduction number is the ratio of the
infection rate to the recovery rate, \(R_0 = \beta/\gamma\).{]}

@@

Even though Figure \ref{fig:different-scales-standard}'s visual analysis
might be able to present more complex properties of the epidemic, one
might wish to understand how these two simulated epidemics' \(R_0\)
values compare. Even though these two simulated epidemics appear
different in these figures ({[}including having different infection
peaks{]}), there is no real intuitive way to compare the epidemic's
\(R_0\). In fact, both of these simulated epidemics started out the same
population (1000 people with 10 infected) and have the same \(R_0\)
value, just with slightly different infection and recovery rates
(\(\beta_1, \gamma_1 = 0.8, 0.4\), and
\(\beta_2, \gamma_2 = .64, .32\)). Even with simple generative models,
this time-dependent visualization tools is not able to be used to
compare a very simple but powerful numerical summary.

@@

The faceted time series plot, like that seen in Figure
\ref{fig:different-scales-standard}, not only fails to allow the
practitioner to compare simulated epidemics' \(R_0\) values but also
presented epidemics' trajectory data so that only the marginal
information for each state could be examined at a given time. A more
holistic approach to visualizing the overall path of our simulated
epidemics would be to examine each epidemic traverses the
three-dimensional \((S(t),I(t),R(t))\) space. Figure 3's left plot does
just that, presenting the trajectories of the simulated epidemics in
this three-dimensional space. {[}For state space models like in our
example, given the constraint that \(S(t) + I(t)+R(t)\) is always equal
to \(N\) (the total population size) we can visual these point in a a
two-dimesional \textit{ternary} plot, as seen in Figure
\ref{fig:different-scales-tern}'s right plot.{]} In this both of Figure
\ref{fig:different-scales-tern}'s plots we an observe that the epidemics
are on the same trajectory. In this simple generative example setting,
{[}this indicates that \textcolor{violet}{the} two epidemics have the
same \(R_0\) value{]}. This can be proven mathematically.

@@

\begin{theorem}\label{thm:sir-scale}
 Let there be two \citet{Kermack1927}'s SIR models ($(S_1(t),I_1(t),R_1(t))_{t\geq 0}$ and $(S_1(s),I_1(s),R_1(s))_{s\geq 0}$), with $(S_1(0),I_1(0),R_1(0))=S_2(0),I_2(0),R_2(0))$  Let both models have the same $R_0$ (aka $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$ and define $a > 0$ such that $\beta_2 = a\beta_1$. [Then for all $t > 0$ there exists an $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.]
\end{theorem}

{[}The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
result from \cite{Harko2014} and is shown in detail in Proof
\ref{proof:thm}. The consequence of Theorem \ref{thm:sir-scale} is that
for two SIR models that have the same initial percent of individuals in
each state and \(R_0\) then for every point on the epidemic path of the
first SIR model can be mapped to a point on the epidemic path of the
second SIR model. In other words, the two epidemics form the same
filamental trajectory.{]}

@@

{[}Through the \(R_0\) example, we see that treating epidemics like
filamental trajectories embedded in a lower dimensional space allows us
better compare the overall structure of the epidemic and see how the
population is directly impacted.{]} In this section we present
time-invariant tools that can be applied to complex epidemics where the
epidemic's generative process is unknown and can have more than three
states. These tools leverage the idea that an epidemic can be viewed as
a trajectory and that many properties of the epidemic are well captured
when we do so. This approach is useful when the epidemic of interest has
only gone {[}through a single realization of its outbreak (before the
population of individuals become susceptible again){]}.

The first set of tools allows a practioner to define distances between
epidemics whose time features don't well align. For completed epidemics,
one way to better examine their properties is to represent their
filamental trajectories as a finite sequence of equally spaced points.
{[}This representation induces a natural distance
\textcolor{violet}{\sout{between this type of representation}} between
epidemics, specifically:{]}

\[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds
\] where \(\tilde{\psi}_i(s)\) the point along \(\psi_i\) that is
\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\). This
distance is naturally time-invariant, and can be plugged into multiple
distance-based assessment tools to examine the overall ``extremeness''
of points, including pseudo-density estimators and depth/local depth
functions \citep[for examples see][]{Ciollaro2016, Geenens2017}. These
extremeness estimators can be useful when comparing the true epidemic to
a a set of simulated epidemics,{]} and practitioners can interpret the
univariate epidemic's extremeness score relative to the extremeness
scores of the simulations very easily.

When the goal is to predict the future impact / trajectory of a
epidemic, time-invariant tools can help describe state-space regions in
which we expect the true epidemic to traverse. In settings where the
epidemic only goes a single outbreak, these regions can be very telling
if simulation models well capture the epidemic's structure. In
\pkg{EpiCompare} we create geometric prediction regions around all but
the \(\alpha\) proportion of most extreme simulated epidemics. These
geometric regions can also be used to compared simulation models that
have different time scales, parameters and even different statistical
philosophies, through set different distances. Although visualization is
the easiest when the epidemic has three states, prediction regions can
be useful to assess and compare simulation models where the epidemics
have multiple states.

@@

\caption{\label{tab:hags-extreme}The extremeness of the true simulations based on comparing psuedo-density estimates between true vs simulated curves}
\centering
\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{6cm}>{\raggedleft\arraybackslash}p{6cm}}
\toprule
Type & simulations-based estimated psuedo-density & proportion of simulations with lower estimated psuedo-density\\
\midrule
Simple SIR & 0.0117621 & 0.01\\
EpiModel SIR & 0.0356503 & 0.01\\
\bottomrule
\end{tabular}
\end{table}

>>>>

EpiCompare emphasizes the value of analyzing epidemics in a
\textit{time-invariant} way - approaches that remove some or all of the
impact of start/end times and recording time scales when preforming the
analysis. In this section we will highlight weaknesses of
time-\textit{de}pendent analysis and define the mathematical
underpinning of the time-invariant approach we take. To accomplish this
goal we will first demonstrate the inability of time-dependent tools to
well quantify a classic epidemic parameter, the reproduction number
\(R_0\). Then we will motivate new time-invariant approaches for more
complex situations where \(R_0\) doesn't capture the complexities of
outbreaks. This will lead us to our final subsection that defines ways
to view epidemics in a time-invariant lense and discusses natural
properties of these representations that allow for clearer ways to
compare models and epidemics.

@@

Epidemiological research is often interested in understanding intrinsic
properties of the epidemic, independent from when the epidemic occurred
or how frequently data was collected. One of the most famous numerical
summaries of an epidemic is the reproduction number - \(R_0\). This
intrinsic
property\footnote{added to shorten this long sentence and connect that it is an intrinsic property}
is a time-invariant value that is defined as the expected number of
infections caused by a single infect or who is added to a completely
susceptible population \citep{anderson1992}. \citet{Gallagher2020}
showed that the estimation of \(R_0\) can be sensitive to time-dependent
parameters like estimation of the beginning and end of an epidemic. This
time-dependent sensitivity in estimating \(R_0\) reflects a general
problem in epidemiology surrounding how to transform time-dependent
information into insights into desirable, intrinsic properties of an
epidemic.

In many situations, epidemiologists want a much deeper understand of an
epidemic than just the reproduction number. One common time-dependent
tool \textcolor{orange}{used to better understand epidemics} is a series
of time series line plots that track the proportion of the population in
a given state (e.g.~infected) at a given time. Figure
\ref{fig:different-scales-standard} visualizes two different simulated
epidemics with population states (S)usceptible, (I)nfectious and
(R)ecovered. These simulations have been generated under a discrete
approximation of \citet{Kermack1927}'s SIR model. The Kermack and
McKendrick model captures the transitions from one state to the next as
a system of ordinary differential equations, where \(N\) is the total
number of individuals, \(\beta\) is the rate of infection, and
\(\gamma\) is the rate of recovery,

@@

From this model, the reproduction number is the ratio of the infection
rate to the recovery rate, \(R_0 = \beta/\gamma\).

@@

Even though Figure \ref{fig:different-scales-standard}'s visual analysis
might be able to present more complex properties of the epidemic, one
might wish to understand how these two simulated epidemics' \(R_0\)
values compare. Even though these two simulated epidemics appear
different in these figures (including having different infection peaks),
there is no real intuitive way to compare the epidemics' \(R_0\) values.
In fact, both of these simulated epidemics started with the same
population (1000 total people with 10 infected) and have the same
\(R_0\) value, just with slightly different infection and recovery rates
(\(\beta_1, \gamma_1 = 0.8, 0.4\), and
\(\beta_2, \gamma_2 = .64, .32\)). Even for simple generative models,
this time-dependent visualization cannot be used to compare a very
simple but powerful numerical summary, \(R_0\).

@@

The faceted time series plot, like that seen in Figure
\ref{fig:different-scales-standard}, not only fails to allow the
practitioner to compare simulated epidemics' \(R_0\) values but also
presented epidemics' trajectory data so that only the marginal
information for each state can be examined at a given time. A more
holistic approach to visualizing the overall path of our simulated
epidemics is to examine how each epidemic traverses the
three-dimensional \((S(t),I(t),R(t))\) space. Figure 3's left plot does
just that, presenting the trajectories of the simulated epidemics in
this three-dimensional space. For state space models like in our
example, given the constraint that \(S(t) + I(t)+R(t)\) is always equal
to \(N\) (the total population size) we can visual these point in a a
two-dimensional \textit{ternary} plot, as seen in Figure
\ref{fig:different-scales-tern}'s right plot. In this both of Figure
\ref{fig:different-scales-tern}'s plots we observe that the two
epidemics are on the same trajectory in this three-dimensional space. In
this simple generative example setting, being on the same trajectory
indicates that \textcolor{violet}{the} two epidemics have the same
\(R_0\) value. This can be proven mathematically.

@@

\begin{theorem}\label{thm:sir-scale}
 Let there be two \citet{Kermack1927}'s SIR models ($(S_1(t),I_1(t),R_1(t))_{t\geq 0}$ and $(S_1(s),I_1(s),R_1(s))_{s\geq 0}$), with $(S_1(0),I_1(0),R_1(0))=S_2(0),I_2(0),R_2(0))$  Let both models have the same $R_0$ (i.e. $R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}$) and define $a > 0$ such that $\beta_2 = a\beta_1$. Then for all $t > 0$ there exists $s>0$ such that $(S_1(t), I_1(t), R_1(t)) = (S_2(s), I_2(s), R_2(s))$.  Moreover, $s = \frac{1}{a}t$.
\end{theorem}

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
result from \cite{Harko2014} and is shown in detail in Proof
\ref{proof:thm}. The consequence of Theorem \ref{thm:sir-scale} is that
for two SIR models that have the same initial percent of individuals in
each state and \(R_0\), then every point on the epidemic path of the
first SIR model can be mapped to a point on the epidemic path of the
second SIR model. In other words, the two epidemics form the same
filamental trajectory.

@@

Through the \(R_0\) example, we see that treating epidemics like
filamental trajectories embedded in a lower dimensional space
(i.e.~through a time-invariant lens) allows us to better compare the
overall structure of the epidemic and see how the population is directly
impacted. In this section we present time-invariant tools that can be
applied to complex epidemics where the epidemic's generative process is
unknown and can have more than three states. These tools leverage the
idea that an epidemic can be viewed as a trajectory and that many
properties of the epidemic are well captured when we do so. This
approach is useful when the epidemic of interest has only gone through a
single realization of its outbreak (before the population of individuals
become susceptible again).

The first set of tools allows a practioner to define distances between
epidemics whose time features do not align. For completed epidemics, one
way to better examine their properties is to represent their filamental
trajectories as a finite sequence of equally spaced points. This
representation induces a natural distance between epidemics,
specifically:

\[
d_\text{equi-distance}(\psi_1, \psi_2)  = \int_{s \in [0,1]} (\tilde{\psi}_1(s) - \tilde{\psi}_2(s))^2 ds
\] where \(\tilde{\psi}_i(s)\) is the point along \(\psi_i\) that is
\(s\cdot|\psi_i|\) distance away from the start of \(\psi_i\), where
\(|\psi_i|\) is the length of \(\psi_i\). This distance is naturally
time-invariant, and can be plugged into multiple distance-based
assessment tools to examine the overall ``extremeness'\,' of points,
including pseudo-density estimators and depth/local depth functions
\citep[for examples see][]{Ciollaro2016, Geenens2017}. These extremeness
estimators can be useful when comparing the true epidemic to a a set of
simulated epidemics, and practitioners can interpret the epidemic's
extremeness score relative to the extremeness scores of the simulations
very easily.

\sout{When the goal is to predict the future impact / trajectory of a epidemic,}
In addition to being used to better understand completed epidemics,
time-invariant tools can be used to aid prediction of future epidemics
by providing a state-space regions in which we expect the true epidemic
to traverse. In settings where the epidemic only generates a single
outbreak, these regions can be very telling if simulation models capture
the epidemic's structure. In \pkg{EpiCompare} we create geometric
prediction regions around all but the \(\alpha\) proportion of most
extreme simulated trajectories. These geometric regions can also be used
to compared simulation models that have different time scales,
parameters and even different statistical philosophies, through set
different distances. Although visualization is the easiest when the
epidemic has three states, prediction regions can be useful to assess
and compare simulation models where the epidemics have multiple states.

@@

\caption{\label{tab:hags-extreme}The extremeness of the true simulations based on comparing psuedo-density estimates between true vs simulated curves}
\centering
\begin{tabular}[t]{l>{\raggedleft\arraybackslash}p{6cm}>{\raggedleft\arraybackslash}p{6cm}}
\toprule
Type & simulations-based estimated psuedo-density & proportion of simulations with lower estimated psuedo-density\\
\midrule
Simple SIR & 0.0036733 & 0.00\\
EpiModel SIR & 0.0149686 & 0.02\\
\bottomrule
\end{tabular}
\end{table}
